---
title: "1. Data Processing" 
author: "Jaeseok Hwang" 
date: "r Sys.Date()" 
output:
 html_document: default 
 pdf_document: default
 word_document: default
---

# PURPOSE

- This R Markdown script executes the data processing phase of the project pipeline.

- Its main objective is to take raw on-farm trial data and integrate it with various external
  geospatial datasets to create a comprehensive, analysis-ready dataset for each field-year.

- The key steps include:
    1. Loading raw experimental data (yield, as-applied rates, field boundaries).
    2. Fetching and processing non-experimental data layers, including:
       - Topography (elevation, slope, aspect) from a DEM.
       - Soil properties (sand, silt, clay content) from the SSURGO database.

    3. Downloading and calculating in-season weather variables (precipitation, GDD, EDD) from Daymet.

    4. Merging all experimental and non-experimental data into a single sf object for each trial.

    5. Saving the final, processed data files, which will serve as the input for the analysis script (2.Data_Analysis.Rmd).


## Knitr option

```{r, cache = F, echo = F, results = "hide"}
#####

library(knitr)

knitr::opts_chunk$set(
  cache = FALSE,
  echo = FALSE,
  warning = FALSE,
  cache.lazy = FALSE,
  fig.retina = 6,
  fig.height = 9,
  fig.width = 9,
  message = FALSE,
  error = TRUE
)

options(knitr.duplicate.label = "allow")

```

#### Packages 

```{r pacakages, cache = FALSE, results = "hide"}

library(here)
library(rmarkdown)
library(jsonlite)
library(parallel)
library(bookdown)
library(knitr)
library(stringr)

library(measurements)
library(data.table)
library(tidyverse)
library(smoother)
library(dplyr)
library(tmap)

library(sf)
library(stars)
library(raster)
library(exactextractr)
library(terra)
library(spatialEco)
library(elevatr)
library(soilDB)
library(FedData)
library(daymetr)
  

```

# Preparation

```{r preparations, cache = T, results = "hide"}

# Read the field parameter data 
field_data <- jsonlite::fromJSON(
 file.path(
    here("Data", "Raw"),
    "field_parameter.json"
  ),
  flatten = TRUE
)%>%
  data.table() 

conv_table <- jsonlite::fromJSON(
 file.path(
    here("Data", "Raw"),
    "nitrogen_conversion.json"
  ),
  flatten = TRUE
)%>%
  data.table() 


# Read functions for data processing 
source(here("Code","Main","0_Set_up.R"))
source(here("Code","Functions","functions_processing.R"))

```

# Data Processing

```{r data processing, echo = F, results = "hide"}

# Read trial fields list ( corn, 102 fields data)

ffy_id_data <- list.files(here("Data","Raw","exp_tb_data")) %>%
 str_subset("_tb.rds") %>%
   str_remove("_tb.rds")

ffy_id_bdry <- list.files(here("Data","Raw","exp_bdry_data")) %>%
 str_subset("_bdry.rds") %>%
   str_remove("_bdry.rds")  



# Check if ffy_id in the trial field list and boundary list are matched
match(ffy_id_data,ffy_id_bdry)

# Read Exp_field_data and process non-exp data on the experimental data
# add weather information on the field parameter ( field specific experimental information)
 
 # "15_1,2,3_2023" (10 ~12)  Out of State 
 # "27_1_2023" (41)
 # "9_1,2_2022" (100~101)

ffy_id_dat <- ffy_id_data[!ffy_id_data %in% c("15_1_2023", "15_2_2023", "15_3_2023", "27_1_2023", "9_1_2022", "9_2_2022")]

saveRDS(ffy_id_dat, here("Data","Processed","ffy_id_list.rds"))

```

```{r stage specified weather info, echo = F, results = "hide"}

# --- Step 3: Load and Filter Initial Data Manifest ---

date_info_raw <- readRDS(here("Data","Raw","date_product_info.rds"))

setDT(date_info_raw)

print(paste("Loaded raw date info with", nrow(date_info_raw), "rows."))

# ** NEW: Filter date_info_raw to only include the valid ffy_ids from ffy_id_dat **
date_info_filtered <- date_info_raw[ffy_id %in% ffy_id_dat]

print(paste("Filtered date info to", nrow(date_info_filtered), "rows."))


# Select and rename the main columns we'll be working with
date_info <- date_info_filtered[, .(
  ffy_id,
  yield_time,
  n_time,
  s_time,
  n_product,
  s_product
)]

# --- Step 4: Process All Fields ---

# Initialize the main list to store all processed data
analysis_list <- list()
cat(sprintf("Starting to process all %d fields from the manifest...\n", nrow(date_info)))

for (i in seq_len(nrow(date_info))) {
  current_ffy_id <- date_info$ffy_id[i]
  timing_info <- copy(date_info[i])
  cat(paste0("\nProcessing: ", current_ffy_id, "\n"))

  # a) Load subplot-level data
  subplot_file_path <- here("Data", "Processed", "Analysis_ready", paste0(current_ffy_id, "_merged_data.rds"))
  if (!file.exists(subplot_file_path)) {
    warning("  - Subplot data file not found. Setting list element to NULL.")
    analysis_list[[current_ffy_id]] <- NULL
    next
  }

  dat_sf <- readRDS(subplot_file_path)
  dat_sf <- st_transform(dat_sf, 4326)
  dat_sf_valid <- st_make_valid(dat_sf)

  field_centroid <- st_centroid(st_make_valid(st_union(dat_sf_valid)))
  centroid_coords <- st_coordinates(field_centroid)

  subplot_centroids <- st_centroid(dat_sf_valid$geom)
  subplot_coords <- st_coordinates(subplot_centroids)

  dat_dt <- as.data.table(st_drop_geometry(dat_sf_valid))
  dat_dt[, `:=`(lon = subplot_coords[, "X"], lat = subplot_coords[, "Y"])]

  timing_info[, `:=`(centroid_lon = centroid_coords[, "X"], centroid_lat = centroid_coords[, "Y"])]

  # b) Conditionally calculate timing features
has_complete_dates <- !is.na(timing_info$s_time) && !is.na(timing_info$n_time) && !is.na(timing_info$yield_time)
has_partial_dates  <-  is.na(timing_info$s_time) && !is.na(timing_info$n_time) && !is.na(timing_info$yield_time)

stage_feature_names <- as.vector(outer(
  c("precip","dry_days","max_dry_spell","heavy_rain_days","gdd","edd"),
  paste0("_S",1:4), paste0)
)

postN_feature_names <- c(
  # N -> harvest
  "precip_N_to_yield","gdd_N_to_yield","edd_N_to_yield",
  "dry_days_N_to_yield","max_dry_spell_N_to_yield","heavy_rain_days_N_to_yield",
  # rolling post-N windows
  paste0("precip_postN_d", c(7,15,30)),
  paste0("heavy_rain_days_postN_d", c(7,15,30)),
  paste0("gdd_postN_d", c(7,15,30)),
  paste0("edd_postN_d", c(7,15,30)),
  # antecedent pre-N windows
  paste0("precip_preN_d", c(7,15,30)),
  paste0("dry_days_preN_d", c(7,15,30)),
  paste0("heavy_rain_days_preN_d", c(7,15,30))
)

nutrient_feature_names <- c("n_app_stage","precip_15_post_N","days_to_N_app")
all_feature_names <- c(stage_feature_names, postN_feature_names, nutrient_feature_names)

final_info <- copy(timing_info)

if (has_complete_dates || has_partial_dates) {
  yr <- if (!is.na(final_info$yield_time)) year(final_info$yield_time) else year(final_info$n_time)

  daily_weather <- tryCatch({
    daymetr::download_daymet(
      lat = final_info$centroid_lat,
      lon = final_info$centroid_lon,
      start = yr, end = yr, internal = TRUE
    )$data
  }, error = function(e) NULL)

  if (is.null(daily_weather)) {
    warning("  - Daymet download failed. Setting timing features to NA.")
    final_info[, (all_feature_names) := NA]
  } else {
    # (A) Post-N features (no s_time needed)
    postN_feats <- compute_postN_features(
      daily_weather,
      final_info$n_time,
      final_info$yield_time
    )

    # (B) Stage features (need s_time); if missing, set NA placeholders
    if (has_complete_dates) {
      stage_feats <- compute_stage_features(
        daily_weather,
        final_info$s_time,
        final_info$n_time,
        final_info$yield_time
      )
    } else {
      stage_feats <- as.list(rep(NA_real_, length(stage_feature_names)))
      names(stage_feats) <- stage_feature_names
      stage_feats <- c(stage_feats, list(
        n_app_stage      = NA_character_,
        precip_15_post_N = NA_real_,
        days_to_N_app    = NA_real_
      ))
    }

    # (C) Legacy 15-day precip after N
    precip_15_post_N_val <- NA_real_
    if (!is.na(final_info$n_time)) {
      n_doy <- yday(final_info$n_time)
      win15 <- as.data.table(daily_weather)[yday > n_doy & yday <= (n_doy + 15)]
      setnames(win15,
               c("prcp..mm.day.","tmax..deg.c.","tmin..deg.c."),
               c("prcp","tmax","tmin"), skip_absent = TRUE)
      precip_15_post_N_val <- sum(win15$prcp, na.rm = TRUE)
    }

    # (D) Merge & assign
    assign_list <- c(
      stage_feats,
      postN_feats,
      list(precip_15_post_N = precip_15_post_N_val)
    )

    # Fill days_to_N_app if s_time exists (stage_feats left NA otherwise)
    if (!is.na(final_info$n_time) && !is.na(final_info$s_time)) {
      assign_list$days_to_N_app <- yday(final_info$n_time) - yday(final_info$s_time)
    }

    for (nm in names(assign_list)) {
      final_info[, (nm) := assign_list[[nm]]]
    }
  }
} else {
  cat("  - Incomplete dates. All timing features will be NA.\n")
  final_info[, (all_feature_names) := NA]
}

# d) Store (don’t forget this before the loop continues)
analysis_list[[current_ffy_id]] <- list(
  info = final_info,
  data = dat_dt
)

}


# --- Step 5: Save the Final, Comprehensive Data Object ---
saveRDS(analysis_list, here("Data", "Processed", "analysis_data_list_full.rds"))

print(paste("\nProcessing complete. Final list with", length(analysis_list), "elements saved."))



```